DDS4Chl: A dataset of drought indices of water demand and supply and its impact on vegetation for continental Chile

### 2024-01-08 article/paper_drought_Chile2023.qmd
last update

In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products. The specific objectives of the study are: i) to calculate multi-scalar drought indices for water demand and supply for 1981–2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

##	Drought indices for water demand and supply

@@

![Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone.

::: {#fig-elephants layout-ncol=1 fig.pos='!ht'}
![](../output/figs/LC_pers80_per_macrozone.png){width="7cm"}

![](../output/figs/table_var_landcover_macro.png){width="8cm"}

Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone. The table on the left shows the linear change trend next to time-series plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021 through the fives zones on continental Chile. Blue dots on the plots indicate the maximum and red dots minimum surface reach.
:::

@@

##	Drought indices for water demand and supply

##	Biomass proxy with drought indices of supply and demand

>>>>

In this research, we aim to analyze the impact of drought on different types of land cover classes in continental Chile by combining environmental variables such as biomass productivity, and water demand and supply gathered from earth observation products. The specific objectives of the study are: i) to analize the trend on multi-scalar drought indices for water demand/supply, soil moisture and vegetation productivity for 1981–2023/2001-2023; ii) to evaluate LULC change for 2001–2021 and its relation to drought indices; iii) to analyze the relationship of a proxy of biomass (zcNDVI) with drought indices; and iv) to assess if the observed changes in the drought indices are linked to TWS.

@@

## Drought indices

##

##	Drought indices for water demand/supply, soil moisture and vegetation productivity

@@

```{r LCprop}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Proportion of land cover class from the persistent land cover for 2001-2021 (>80%) per macrozone"
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/LC_pers80_per_macrozone.png")
```

@@

##	Trend of drought indices for water demand/supply, soil moisture, and vegetation productivity

Regarding vegetation productivity, there is a negative trend on zcNDVI in the central part of the country ("norte chico" and "zona central") (@fig-zcNDVI_var). Indicating either a reduction of vegetation surface, biomass, or status. It reached its lowest values during the year 2020, mostly on the "zona central." There are other negative trends in small areas in "norte grande" and "zona asutral."

```{r fig-zcNDVI_var}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "(a) Map of the linear trend of the index zcNDVI-6 for 2001–2023. Greener colors indicate a positive trend and an increase in vegetation development; reder colors correspond to a negative trend and a decrease in vegetation development. (b) Temporal variation of zcNDVI-6 aggregated at macrozone level within continental Chile. Each horizontal panel corresponds to a macrozone from 'norte grande' to 'zona austral'."
#| out-width: "\\textwidth"
knitr::include_graphics("../output/figs/temporal_variation_zcNDVI6_macrozonas.png")
```

Analyzing the water supply in the macrozones "norte chico," "central," and "sur," the SPI shows a decreasing trend that increases at longer time scales due to the prolonged reduction in precipitation. It reaches a trend of -0.056 (z-score) per decade (@fig-trendSPI) for time scales of 36 months. In the case of water demand, we analyze the EDDI. It shows a positive trend, caused by an increase in AED. The trend on EDDI reaches a maximum of 0.053 per decade in the macrozones "norte grande" and "norte chico" (@fig-trendEDDI). The behavior is similar to SPI, having close trends with opposite signs. Nevertheless, the spatial patterns are different. The maximum trend for SPI-36 is in "norte chico" and "zona central." For EDDI-36, the maximum trend is from "northe grande" to "zona central."

```{r fig-trendSPI}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Linear trend of the SPI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
#| out-width: "\\textwidth"

knitr::include_graphics("../output/figs/trend_raster_SPI_1981-2023.png")
```

```{r fig-trendEDDI}
#| echo: false
#| fig-pos: "!ht"
#| out-width: "\\textwidth"
#| fig-cap: "Linear trend of the EDDI at time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023"
knitr::include_graphics("../output/figs/trend_raster_EDDI_1981-2023.png")
```

The correlation between zcNDVI-6 and drought indices shows that time scales higher than six months are predominant over continental Chile in explaining vegetation variability (@fig-corTimeScale). Pearson correlations between 0.6 and 0.9 are concentrated in the "norte chico" and "zona central" for SPI and SPEI. Soil moisture reached a correlation of 0.6–0.9 that extended north and south with respect to SPI.

```{r fig-corTimeScale}
#| echo: false
#| fig-pos: "!ht"
#| out-width: "\\textwidth"
#| fig-cap: "Time scales per drought index that reach the maximum coefficient of determination"
knitr::include_graphics("../output/figs/mapa_cor_selec_indices_zcNDVI6.png")
```

```{r fig-corPerson}
#| echo: false
#| fig-pos: "!ht"
#| out-width: "\\textwidth"
#| fig-cap: "Pearson correlation value for the time scales and drought index that reach the maximum coefficient of determination"
knitr::include_graphics("../output/figs/mapa_cor_r_indices_zcNDVI6.png")
```

### 2024-01-08 article/paper_drought_Chile2023.tex
last update

\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from the ERA5-Land and MODIS},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Assessment of drought impact over land cover in continental Chile
by the analysis of water supply and demand from the ERA5-Land and MODIS}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt, breakable, enhanced]}{\end{tcolorbox}}\fi

@@

In this research, we aim to analyze the impact of drought on different
types of land cover classes in continental Chile by combining
environmental variables such as biomass productivity, and water demand
and supply gathered from earth observation products. The specific
objectives of the study are: i) to calculate multi-scalar drought
indices for water demand and supply for 1981--2023; ii) to evaluate LULC
change for 2001--2021 and its relation to drought indices; iii) to
analyze the relationship of a proxy of biomass (zcNDVI) with drought
indices; and iv) to assess if the observed changes in the drought
indices are linked to TWS.

@@

\begin{figure}[ht]

{\centering \includegraphics[width=0.8\textwidth,height=\textheight]{../output/figs/map_study_con_landcover.png}

@@

\hypertarget{drought-indices-for-water-demand-and-supply}{%
\subsection{Drought indices for water demand and
supply}\label{drought-indices-for-water-demand-and-supply}}

@@

!{[}Proportion of land cover class from the persistent land cover for
2001-2021 (\textgreater80\%) per macrozone.

\begin{figure}

\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics[width=7cm,height=\textheight]{../output/figs/LC_pers80_per_macrozone.png}

@@

\end{minipage}%
\newline
\begin{minipage}[t]{\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics[width=8cm,height=\textheight]{../output/figs/table_var_landcover_macro.png}

\end{minipage}%

\caption{\label{fig-elephants}Proportion of land cover class from the
persistent land cover for 2001-2021 (\textgreater80\%) per macrozone.
The table on the left shows the linear change trend next to time-series
plot of surface, per land cover class (IGBP MCD12Q1.006) for 2001-2021
through the fives zones on continental Chile. Blue dots on the plots
indicate the maximum and red dots minimum surface reach.}

@@

In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
solutions were used, which are provided by the Jet Propulsion Laboratory
(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
0.5 degrees represents the deviation of surface mass for that month
relative to a reference time average (2004-2009 baseline), which is
subtracted from all other monthly grids to provide terrestrial water
storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
version of the data employs a Coastal Resolution Improvement (CRI)
filter that reduces signal leakage errors across coastlines
\citep{Wiese2019, Wiese2016}. Although the mascon solutions greatly
reduced leakage errors, a gain factor, which is used to enhance the
spatial resolution, was applied to the dataset as recommended for
hydrological studies \citep{Ramjeawon2022, Yin2022}. The water storage
and height anomalies are given in Equivalent Water Height or Thickness
units (EWH, cm) \citep{Sabzehee2023}, and their temporal resolution
monthly is from April 2002 to May 2023.

@@

\hypertarget{drought-indices-for-water-demand-and-supply-1}{%
\subsection{Drought indices for water demand and
supply}\label{drought-indices-for-water-demand-and-supply-1}}

\hypertarget{biomass-proxy-with-drought-indices-of-supply-and-demand}{%
\subsection{Biomass proxy with drought indices of supply and
demand}\label{biomass-proxy-with-drought-indices-of-supply-and-demand}}

>>>>

\usepackage{lineno}\linenumbers
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Assessment of drought impact over land cover in continental Chile by the analysis of water supply and demand from ERA5-Land and MODIS},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Assessment of drought impact over land cover in continental Chile
by the analysis of water supply and demand from ERA5-Land and MODIS}
\author[1]{Francisco Zambrano%
\corref{cor1}%
\fnref{fn1}}
 \ead{francisco.zambrano@umayor.cl}

@@

\begin{keyword}
    drought \sep land cover change \sep 
    satellite
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, boxrule=0pt, interior hidden, breakable]}{\end{tcolorbox}}\fi

@@

In this research, we aim to analyze the impact of drought on different
types of land cover classes in continental Chile by combining
environmental variables such as biomass productivity, and water demand
and supply gathered from earth observation products. The specific
objectives of the study are: i) to analize the trend on multi-scalar
drought indices for water demand/supply, soil moisture and vegetation
productivity for 1981--2023/2001-2023; ii) to evaluate LULC change for
2001--2021 and its relation to drought indices; iii) to analyze the
relationship of a proxy of biomass (zcNDVI) with drought indices; and
iv) to assess if the observed changes in the drought indices are linked
to TWS.

@@

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/map_study_con_landcover.png}

@@

\hypertarget{drought-indices}{%
\subsection{Drought indices}\label{drought-indices}}

\hypertarget{section}{%
\subsection{}\label{section}}

\hypertarget{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
\subsection{Drought indices for water demand/supply, soil moisture and
vegetation
productivity}\label{drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}

@@

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/LC_pers80_per_macrozone.png}

@@

\caption{Proportion of land cover class from the persistent land cover
for 2001-2021 (\textgreater80\%) per macrozone}

@@

In this study, RL06.1\_V3 GRACE mascon (mass concentration) monthly
solutions were used, which are provided by the Jet Propulsion Laboratory
(JPL-M, https://grace.jpl.nasa.gov). Each GRACE Tellus monthly grid at
0.5 degrees represents the deviation of surface mass for that month
relative to a reference time average (2004-2009 baseline), which is
subtracted from all other monthly grids to provide terrestrial water
storage anomalies (TWSA) \citep{Ramjeawon2022, Yin2022}. This JPL-M
version of the data employs a Coastal Resolution Improvement (CRI)
filter that reduces signal leakage errors across coastlines
\citep{NASA2019, Wiese2016}. Although the mascon solutions greatly
reduced leakage errors, a gain factor, which is used to enhance the
spatial resolution, was applied to the dataset as recommended for
hydrological studies \citep{Ramjeawon2022, Yin2022}. The water storage
and height anomalies are given in Equivalent Water Height or Thickness
units (EWH, cm) \citep{Sabzehee2023}, and their temporal resolution
monthly is from April 2002 to May 2023.

@@

\hypertarget{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}{%
\subsection{Trend of drought indices for water demand/supply, soil
moisture, and vegetation
productivity}\label{trend-of-drought-indices-for-water-demandsupply-soil-moisture-and-vegetation-productivity}}

Regarding vegetation productivity, there is a negative trend on zcNDVI
in the central part of the country (``norte chico'' and ``zona
central'') (Figure~\ref{fig-zcNDVI_var}). Indicating either a reduction
of vegetation surface, biomass, or status. It reached its lowest values
during the year 2020, mostly on the ``zona central.'' There are other
negative trends in small areas in ``norte grande'' and ``zona asutral.''

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/temporal_variation_zcNDVI6_macrozonas.png}

\caption{\label{fig-zcNDVI_var}(a) Map of the linear trend of the index
zcNDVI-6 for 2001--2023. Greener colors indicate a positive trend and an
increase in vegetation development; reder colors correspond to a
negative trend and a decrease in vegetation development. (b) Temporal
variation of zcNDVI-6 aggregated at macrozone level within continental
Chile. Each horizontal panel corresponds to a macrozone from `norte
grande' to `zona austral'.}

Analyzing the water supply in the macrozones ``norte chico,''
``central,'' and ``sur,'' the SPI shows a decreasing trend that
increases at longer time scales due to the prolonged reduction in
precipitation. It reaches a trend of -0.056 (z-score) per decade
(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
water demand, we analyze the EDDI. It shows a positive trend, caused by
an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
decade in the macrozones ``norte grande'' and ``norte chico''
(Figure~\ref{fig-trendEDDI}). The behavior is similar to SPI, having
close trends with opposite signs. Nevertheless, the spatial patterns are
different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
``zona central.''

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/trend_raster_SPI_1981-2023.png}

\caption{\label{fig-trendSPI}Linear trend of the SPI at time scales of
1, 3, 6, 12, 24, and 36 months for 1981-2023}

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/trend_raster_EDDI_1981-2023.png}

\caption{\label{fig-trendEDDI}Linear trend of the EDDI at time scales of
1, 3, 6, 12, 24, and 36 months for 1981-2023}

The correlation between zcNDVI-6 and drought indices shows that time
scales higher than six months are predominant over continental Chile in
explaining vegetation variability (Figure~\ref{fig-corTimeScale}).
Pearson correlations between 0.6 and 0.9 are concentrated in the ``norte
chico'' and ``zona central'' for SPI and SPEI. Soil moisture reached a
correlation of 0.6--0.9 that extended north and south with respect to
SPI.

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/mapa_cor_selec_indices_zcNDVI6.png}

\caption{\label{fig-corTimeScale}Time scales per drought index that
reach the maximum coefficient of determination}

\begin{figure}[!ht]

{\centering \includegraphics{../output/figs/mapa_cor_r_indices_zcNDVI6.png}

\caption{\label{fig-corPerson}Pearson correlation value for the time
scales and drought index that reach the maximum coefficient of
determination}

### 2024-01-08 conference/IGARSS2024.qmd
last update

note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
#keywords: [IEEE, IEEEtran, journal, Quarto, Pandoc, template]
#funding: 
funding: 
  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022 and 11190360."
pageheader:
  left: IGARSS, January 2024
  right: 'D. Folio:  A Sample Article Using quarto-ieee'
bibliography: references.bib
date: 2024-01-05
pdf: https://github.com/dfolio/quarto-ieee/blob/main/template.pdf
citation: 
  container-title: GitHUB
  page: 1-3
  type: software
  issued: 2023-06-23
  url: https://github.com/dfolio/quarto-ieee
  pdf-url: https://github.com/dfolio/quarto-ieee/template.pdf
---

>>>>

note: "Template created June 23, 2023; revised `r format(Sys.Date(),format='%B %d, %Y')`."
#keywords: [IEEE, IEEEtran, journal, Quarto, Pandoc, template]
#funding: 
funding: 
  statement: "The research was funded by the National Research and Development Agency of Chile (ANID) and the projects FSEQ210022 and 11190360."
pageheader:
  left: IGARSS, January 2024
  right: 'D. Folio:  A Sample Article Using quarto-ieee'
bibliography: references.bib
date: 2024-01-05
---
